---
title: "Fundamentos de Programación y Visualización de Datos en R"
subtitle: "Introducción a Tidyverse"
author: "Guillermo Basulto-Elias"
date: "2022-11-10"
institute: "InTrans - Iowa State University"
format: revealjs
editor: visual
slide-number: true
---

## Introduction to Tidyverse

-   Los paquetes de R es lo que lo hacen resaltar sobre otros lenguajes de programación

-   Cualquiera puede contribuir con un paquete en cualquier momento

-   Algunos paquetes populares:

    -   **lme4**: Modelos lineales con effectos mixtos.

    -   **Rcpp**: Acelerar código de R con C++.

    -   **caret:** Herramientas de *machine learning*.

    -   **shiny**: Crear tableros interactivos.

## Ventajas y Desventajas de Paquetes de R

-   Ventajas:

    -   Accesibilidad inmediata produce retroalimentación (a través de [CRAN](https://cran.r-project.org/), GitHub, o [Bioconductor](https://www.bioconductor.org/)).

    -   Nuevos métodos están disponibles al instante.

-   Desventajas:

    -   Falta de consistencia entre paquetes y métodos

    -   Los paquetes pueden ser retirados en cualquier instante

    -   Creadores pueden dejar de dar mantenimiento a paq.

    -   Algunos paquetes podrían no ser confiables

## ¿Qué es el Tidyverse?

> The tidyverse is an opinionated [**collection of R packages**](https://www.tidyverse.org/packages) designed for data science. All packages share an underlying design philosophy, grammar, and data structures.

-   Patrocinado por [Posit](https://posit.co/) (formerly RStudio).

-   Ventajas:

    -   Desarrollo constante

    -   Consistencia

    -   Documentation amplia y detallada

## Paquetes Nucleares de Tidyverse

-   Leer archivos

    -   **readr** Archivos de texto
    -   **readxl** Archivos de Excel

-   Gráficos

    -   **ggplot2** Gráficos. Muchas [extensiones útiles](https://exts.ggplot2.tidyverse.org/gallery/).

## Core Tidyverse Packages (cont'd)

-   Preparación/manipulación de datos

    -   **dplyr** Data manipulation centered around data frames.

    -   **tidyr** Helper functions to get tidy data.

    -   **purrr** Toolkit to handle vectors and lists.

    -   **tibble** Data frames with attributes to make their content more explicit.

    -   **stringr** Tools to work with strings.

    -   **forcats** Tools to manipulate factors.

    -   **lubridate** (to be included) Date and time operations.

## Packages Loaded by Tidyverse

```{r, eval = TRUE, echo=TRUE, message=TRUE}
## Packages loaded by tidyverse. Pay attention to message.
library(tidyverse)
```

## Alternatives to Tidyverse Packages

-   In R, there might be more than one package doing the same or very similar things.

-   Some alternatives:

    -   **ggplot2** `lattice`, base plots.

    -   **Data manipulation** `data.table`, apply family (`apply`, `sapply`, `mapply`, etc.).

    -   **Read files** `data.table`, base funcions (`read.csv`, `read.delim`, etc.)

## Main Resources

-   [Website of each package](https://www.tidyverse.org/packages/) in tidyverse website.

-   [RStudio Cheatsheets](https://www.rstudio.com/resources/cheatsheets/). Includes others (e.g., shiny, reticulate, Keras, Spark).

-   [R for Data Science](https://r4ds.had.co.nz/) (available online).

## Website of a Tidyverse Package

![Structure of [package website](https://dplyr.tidyverse.org/articles/dplyr.html).](images/website_screenshot.png)

## tibble

-   Data frames with attributes to make their content more explicit.

-   Prints nicer data frames.

-   Column type is clearly visible.

-   Main functions:

    -   `tibble()` Similar to data.frame, but returns a tibble object.
    -   `as_tibble()` Casts a data frame to tibble by adding attributes
    -   `tribble()` Creates small data frame.

## readr

-   Fast way to read data.
-   Main functions:
    -   `read_csv()` Reads comma-delimited files.
    -   `read_delim()` Reads delimited files.
-   Others:
    -   `cols()` Specifies column types.
    -   `parse_double()`, `parse_logical()`, etc. Parse specific types.

## dplyr

-   Data manipulation centered around **data frames**.

-   Main functions:

    -   `mutate()` Modifies or adds column (same rows).
    -   `select()` Keeps specified columns only.
    -   `filter()` Subsets rows according to a condition (or more than one).
    -   `summarize()` Summarizes with the provided function (fewer rows)
    -   `arrange()` Sorts data frame according to provided variables.

## dplyr (cont'd)

-   Check the [documentation instead](https://dplyr.tidyverse.org/reference/index.html). Examples by category:

    -   One table verb. `arrange()`, `distinct()`, ...

    -   Two table verbs. `bind_rows()`, `left_join()`, ...

    -   Grouping. `group_by()`, `ungroup()`, ...

    -   Vector functions. `across()`, `case_when()`, ...

    -   Data. `starwars`, `storms`, ...

    -   Remote table. `sql()`, `collect()`, ...

    -   Experimental. `rows_update()`, `rows_path()`, ...

## tidyr

-   Helper functions to get tidy data. See [documentation](https://tidyr.tidyverse.org/reference/index.html).

-   Main functions:

    -   `pivot_longer()` Pivots data from wide to long.
    -   `pivot_wider()` Pivots data from long to wide.

-   Other:

    -   `separate()` Separates a column into two or more cols.
    -   `unite()` Unites multiple columns.

## stringr

-   Tools to work with strings.
-   Ideal to combine with [regular expressions](https://stringr.tidyverse.org/articles/regular-expressions.html).
-   Some functions:
    -   `str_replace_all()` Replaces string or regular expression.

    -   `str_detect()` Detects string or regular expression is present.

    -   `str_length()` Returns number of characters.

## forcats

-   Tools to manipulate factors.

-   A factor is used to categorize a variable with a limited number of values (e.g., day/night, tall/grande/venti, etc.)

-   Required for statistical models and plots

-   Some functions:

    -   `fct_reorder()`, `fct_relevel()`, `fct_collapse()`.

## lubridate

-   Date and time operations. Examples:

-   Date/time parsing: `ymd()`, `mdy()`, `ym()`, `ymd_hms()`.

-   Setting, gettting, rounding: `month()`, `week()`, `day()`.

-   Date helpers: `as_date()`, `is.Date()`, `today()`, `now()`.

-   Period: `seconds()`, `days()`, `months()`.

-   Other function categories: intervals, duration, timespans, formatters, others.

## purrr

-   Consistent family of functions to apply to lists and vectors.

-   A list is a collection of R objects without a set structure, unlike data frames, arrays, matrices or vectors.

    ```{r, eval=TRUE, echo = TRUE}
    ## Example of list with a vector and a matrix.
    mylist <- list(a_vector = 1:3, a_matrix = matrix(c(1, 2, 3, 4), 2))
    mylist
    ```

-   Provides excellent substitutes for the apply family (`apply()`, `mapply()`, `tapply()`, `sapply()`, etc.).

## ggplot2

-   Grammar-of-graphics plots. Plenty of [useful extensions](https://exts.ggplot2.tidyverse.org/gallery/).

-   See [documentation](https://ggplot2.tidyverse.org/).

-   Examples have been covered in class and will be an important part of it.

## More of Tidyverse

-   `readxl` Read Excel documents. Similar to readr.

-   `googlesheets4` Google Sheets.

-   `rvest` Web-scrapping.

-   `jsonlite` JSON.

-   `hms` for time-of-day values.

-   `dbplyr` to convert dplyr into SQL.

-   `tidymodels` A whole new universe to adjust consistent statistical and ML models. Like a tidyverse focus on models.

## Tidyverse-friendly Packages

-   [`fs`](https://www.tidyverse.org/blog/2018/01/fs-1.0.0/) Manage files: Set paths, add, delete or copy files.[^1]

-   [`sf`](https://r-spatial.github.io/sf/) For shapefiles, GeoJSON and similar.

-   [`here`](https://here.r-lib.org/) To unambiguously define your work. Ideal to work on teams via Box, Dropbox, or on different equipments.

-   [`glue`](https://glue.tidyverse.org/) To concatenate strings (paste0(), but better and more flexible).[^2]

-   [`janitor`](https://sfirke.github.io/janitor/) Clean column names and cell content.

[^1]: Just realized it is also part of the Tidyverse.

[^2]: Just realized it is also part of the Tidyverse.

## Questions

Guillermo Basulto-Elias

basulto\@iastate.edu
